---
title: "Strongyloides spp. Gene Annotations"
author: "Astra Bryant"
date: "2024-12-11"
output:
  html_document:
    df_print: paged
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
The goal of this file is to import and clean gene annotation information for all genes from the following *Strongyloides*  species: *S. stercoralis* (SSTP annotations),*S. ratti*, *S. papillosus*, and *S. venezuelensis*.

Annotation information includes:
  * Gene name
  * WormBase gene name/IDs
  * WormBase description
  * InterPro description
  * Homologs/Percent homology with other *Caenorhabditis* species  

```{r setup, echo = FALSE, results = FALSE, include = FALSE}
if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!require("tximport", quietly = TRUE)) BiocManager::install("tximport", ask = FALSE)
if (!require("ensembldb", quietly = TRUE)) BiocManager::install("ensembldb", ask = FALSE)
if (!require("biomaRt", quietly = TRUE)) BiocManager::install("biomaRt", ask = FALSE)
if (!require("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load("tidyverse","data.table", "magrittr","ensembldb", "biomaRt", "knitr")

```

```{r geneAnnotation}
# Introduction to this chunk -----------
# This chunk imports gene annotation information using WBPS BiomaRT
# It will generate and save a table.

# Get gene information from BiomaRT and filter -----
species_list <- tibble(species = c('ststerprjeb528', 'strattprjeb125', 'stpapiprjeb525', 'stveneprjeb530'))
comparison_list <- tibble(species = c('ststerprjeb528', 'strattprjeb125', 'stpapiprjeb525', 'stveneprjeb530', 'caelegprjna13758'))

for (x in species_list$species) {
  comparison.attributes <- dplyr::filter(comparison_list, species!= x ) %>%
  dplyr::mutate(geneID = paste0(species, "_gene"), geneName = paste0(species, "_gene_name"), percID = paste0(species, "_homolog_perc_id"))

  
annotations <- getBM(attributes=c('wbps_transcript_id',
                                  'wbps_gene_id',
                                  'chromosome_name',
                                  'description',
                                comparison.attributes$geneID[[1]],
                                comparison.attributes$geneName[[1]],
                                comparison.attributes$percID[[1]],
                                comparison.attributes$geneID[[2]],
                                comparison.attributes$geneName[[2]],
                                comparison.attributes$percID[[2]],
                                comparison.attributes$geneID[[3]],
                                comparison.attributes$geneName[[3]],
                                comparison.attributes$percID[[3]],
                                comparison.attributes$geneID[[4]],
                                comparison.attributes$geneName[[4]],
                                comparison.attributes$percID[[4]]),
                     # grab the annotations from WormBase ParaSite
                     mart = useMart(biomart="parasite_mart", 
                                    dataset = "wbps_gene", 
                                    host="https://release-18.parasite.wormbase.org", #Using the archived version of biomart because the XLOC gene ids are not useful.
                                    port = 443),
                     filters = 'species_id_1010',
                     useCache = T,
                     value = x) %>%
    as_tibble(.name_repair = "unique") %>%
    dplyr::rename(Description = "Gene description",
                  Location = "Chromosome/scaffold name",
                GS1_percent_homology = "% identity...7",
                GS2_percent_homology = "% identity...10",
                GS3_percent_homology = "% identity...13",
                GS4_percent_homology = "% identity...16"
  ) %>%
    tidyr::unite("GS4_homologID", 15, 14, sep = " | ", remove = T)%>%
    tidyr::unite("GS3_homologID", 12, 11, sep = " | ", remove = T)%>%
    tidyr::unite("GS2_homologID", 9, 8, sep = " | ", remove = T) %>%
    tidyr::unite("GS1_homologID", 6, 5, sep = " | ", remove = T)

annotations$GS1_homologID<- str_remove_all(annotations$GS1_homologID, "^ \\| ")
annotations$GS2_homologID<- str_remove_all(annotations$GS2_homologID, "^ \\| ")
annotations$GS3_homologID<- str_remove_all(annotations$GS3_homologID, "^ \\| ")
annotations$GS4_homologID<- str_remove_all(annotations$GS4_homologID, "^ \\| ")

if (x == "strattprjeb125"){
    annotations <- dplyr::rename(annotations, geneID = "Transcript stable ID", WBgeneID = "Gene stable ID") %>%
        group_by(geneID)
    annotations$geneID <- str_remove_all(annotations$geneID, "\\.[0-9]$")
    annotations$geneID  <- str_remove_all(annotations$geneID, "[a-z]$")
} else {
    annotations <- dplyr::rename(annotations, geneID = "Gene stable ID") %>%
        dplyr::select(!"Transcript stable ID")
    
}

if (x == "ststerprjeb528"){
    chromosomes <- read.table(file = "./Data/Ss/SSTP_XLOC_overlapping_genes.txt",
                              header = T) %>%
        tidyr::unite("XLOC_IDs", 2, 3, sep = " | ", remove = T)
    annotations <- dplyr::select(annotations, !"Location") %>%
        left_join(chromosomes, join_by(geneID == SSTP_GeneID), relationship = "many-to-many") %>%
        dplyr::rename(Location = Chromosome) %>%
        dplyr::relocate(XLOC_IDs:Location, .after = geneID)
} 

# Gather then clean interpro description, removing duplications and collapsing to one line
annt.temp.3 <- getBM(attributes=c('wbps_gene_id', 
                                     'interpro_short_description',
                                     'go_name_1006',
                                     'uniprot_sptrembl'),
                     # grab the annotations from WormBase ParaSite
                     mart = useMart(biomart="parasite_mart", 
                                    dataset = "wbps_gene", 
                                    host="https://release-18.parasite.wormbase.org", #Using the archived version of biomart because the XLOC gene ids are not useful.
                                    port = 443),
                     filters = 'species_id_1010',
                     useCache = T,
                     value = x) %>%
  dplyr::rename(geneID = "wbps_gene_id",
                InterPro = "interpro_short_description",
                GO_term = "go_name_1006",
                UniProtKB = "uniprot_sptrembl") %>%
  group_by(geneID)

# Clean up InterPro, GO, and UniProtKB terms, removing duplications and collapsing to one line each
interpro_terms<-annt.temp.3 %>%
    dplyr::select(geneID, InterPro) %>%
    group_by(geneID) %>%
    subset(InterPro != "") %>%
    dplyr::distinct(InterPro, .keep_all = TRUE) %>%
    dplyr::summarise(InterPro = paste(InterPro, 
                                    collapse = ', ')) 
go_terms<-annt.temp.3 %>%
    dplyr::select(geneID, GO_term) %>%
    group_by(geneID) %>%
     subset(GO_term != "") %>%
    dplyr::distinct(GO_term, .keep_all = TRUE) %>%
    dplyr::summarise(GO_term = paste(GO_term, collapse = ', '))

UniProtKB_terms<-annt.temp.3 %>%
    dplyr::select(geneID, UniProtKB) %>%
    group_by(geneID) %>%
     subset(UniProtKB != "") %>%
    dplyr::distinct(UniProtKB, .keep_all = TRUE) %>%
    dplyr::summarise(UniProtKB = paste(UniProtKB, collapse = ', '))

if (any(annotations$geneID %in% interpro_terms$geneID)){
annotations <-left_join(annotations, interpro_terms) %>%
    left_join(go_terms) %>%
    left_join(UniProtKB_terms) %>%
  relocate(InterPro, GO_term, UniProtKB, .after = Description) %>%
    group_by(geneID)
} else {
annotations <-left_join(annotations, interpro_terms, by = join_by(WBgeneID == geneID)) %>%
    left_join(go_terms, by = join_by(WBgeneID == geneID)) %>%
    left_join(UniProtKB_terms, by = join_by(WBgeneID == geneID)) %>%
  relocate(InterPro, GO_term, UniProtKB, .after = Description) %>%
    group_by(geneID)
}

# Remove source code to shorten the description
annotations$Description <- annotations$Description %>%
  str_replace_all(string = ., 
                  pattern = "  \\[Source:.*\\]", 
                  replacement = "")

# Replace empty string values with NAs
annotations[annotations == "NA | NA"]<-NA
annotations[annotations == " | "]<-NA

# Remove any duplicates in the possible homolog matches. 
# Select based on highest % homology.
# Give fake value here to make sure genes 
# without homologs aren't filtered out
annotations$GS1_percent_homology[
  is.na(annotations$GS1_percent_homology)] <- 1000 
annotations$GS2_percent_homology[
  is.na(annotations$GS2_percent_homology)] <- 1000
annotations$GS3_percent_homology[
  is.na(annotations$GS3_percent_homology)] <- 1000
annotations$GS4_percent_homology[
  is.na(annotations$GS4_percent_homology)] <- 1000

annotations <-annotations %>%
  slice_max(n = 1, order_by = GS1_percent_homology, 
            with_ties = FALSE) %>%
  slice_max(n = 1, order_by = GS2_percent_homology, 
            with_ties = FALSE) %>%
  slice_max(n = 1, order_by = GS3_percent_homology, 
            with_ties = FALSE)%>%
  slice_max(n = 1, order_by = GS4_percent_homology, 
            with_ties = FALSE)

annotations$GS1_percent_homology[
  annotations$GS1_percent_homology == 1000] <- NA
annotations$GS2_percent_homology[
  annotations$GS2_percent_homology == 1000]<- NA
annotations$GS3_percent_homology[
  annotations$GS3_percent_homology == 1000]<- NA
annotations$GS4_percent_homology[
  annotations$GS4_percent_homology == 1000]<- NA

# Check for presence of output folder, generate if it doesn't exist
output.path <- "./Data"

species <- switch(x,
                  `ststerprjeb528` = 'Ss',
                  `strattprjeb125` = 'Sr',
                  `stpapiprjeb525` = 'Sp',
                  `stveneprjeb530` = "Sv")

app.path <-"../Data"
output.name <- paste0(species, '_geneAnnotations')

save(annotations,
     file = file.path(output.path,
                      species,
                      output.name))

# Save in Shiny app Data folder
save(annotations,
     file = file.path(app.path,
                      output.name))
}

```
